quicR:\ An R Library for Streamlined Data Handling of Real-Time Quaking Induced Conversion Assays

### 2025-02-26 man/manuscript/softwareX_submission.tex
Updated manuscript.

\documentclass[preprint,12pt, a4paper]{elsarticle}

@@

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
% \usepackage{lineno}

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{quicR: An R Library for Streamlined Data Handling of Real-Time Quaking Induced Conversion Assays}

\author[label1,label2,label3]{Gage Rowden\corref{cor}}
\fntext[ ]{\textit{E-mail address: }rowde002@umn.edu}
\cortext[cor]{Corresponding author.}
\author[label1,label2,label3]{Peter Larsen}
\address[label1]{Department of Veterinary and Biomedical Sciences, University of Minnesota, USA.}
\address[label2]{Minnesota Center for Prion Research and Outreach, University of Minnesota, USA.}
\address[label3]{Priogen Corp., USA.}

\begin{abstract}
Real-time quaking induced conversion (RT-QuIC) has quickly become a valuable diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is relatively new, academic and industry standards for quality filtering data and high throughput analysis of results have yet to be fully established. The open source R library, quicR, was developed to provide a standardized approach to RT-QuIC data analysis. quicR provides functions, which can be easily integrated into existing R workflows, for data curation, analysis, and vizualization.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
R package \sep RT-QuIC \sep prion \sep diagnostics \sep CJD \sep Parkinson's

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

@@

\begin{table}[ht]
    \fontsize{9}{9}\selectfont
    \begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
        \hline
        \textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata} \\
        \hline
        C1 & Current code version & V2.1.0 \\
        \hline
        C2 & Permanent link to code/repository used for this code version & \url{https://github.com/gage1145/quicR} \\
        \hline
        C3  & Permanent link to Reproducible Capsule & \url{https://github.com/gage1145/quicR/releases/tag/v2.1.0}\\
        \hline
        C4 & Legal Code License & GPL-3 \\
        \hline
        C5 & Code versioning system used & git \\
        \hline
        C6 & Software code languages, tools, and services used & R \\
        \hline
        C7 & Compilation requirements, operating environments \& dependencies & R (>=4.1.0) \\
        \hline
        C8 & If available Link to developer documentation/manual & \url{https://cran.r-project.org/web/packages/quicR/quicR.pdf} \\
        \hline
        C9 & Support email for questions & rowde002@umn.edu\\
        \hline
    \end{tabular}
\caption{Code metadata}
\label{codeMetadata} 
\end{table}

@@

    By combining standardized metrics, advanced visualization tools, and a commitment to open source science, quicR serves as a foundational resource for the growing RT-QuIC community. Its goal is to empower researchers to analyze and present their data with clarity, consistency, and cohesion.

@@

\section{Software description}
    \subsection{Software architecture}
        quicR was developed to address the growing need for efficient data conversion, analysis, and visualization of RT-QuIC data. With a focus on usability and reproducibility, the package is designed to standardize workflows and ensure compatibility across multiple laboratories. Its primary input format is data exported as Excel workbooks from the proprietary MARS software (BMG Labtech, Ortenberg, Germany), providing seamless integration with existing experimental workflows. See Figure \ref{fig:workflow} for a detailed workflow.
        \begin{figure}[ht]
            \caption{Workflow hierarchy of the quicR package. Blue nodes indicate steps where BMG software is needed. Purple nodes indicate functions dedicated to handling metadata. Red nodes are functions that acquire and manipulate raw data. Orange nodes are functions which calculate some metric. Finally, yellow nodes represent data analysis endpoints.}
            \centering
            \includegraphics[width=\textwidth]{images/workflow2.png}
            \label{fig:workflow}
        \end{figure}

\subsection{Software functionalities}
\textit{  Present the major functionalities of the software.}

 \subsection{Sample code snippets analysis (optional)}

\section{Illustrative examples}

@@

% \textit{\textbf{Optional}: you may include one explanatory  video or screencast that will appear next to your article, in the right hand side panel. Please upload any video as a single supplementary file with your article. Only one MP4 formatted, with 150MB maximum size, video is possible per article. Recommended video dimensions are 640 x 480 at a maximum of 30 frames / second. Prior to submission please test and validate your .mp4 file at  \url{http://elsevier-apps.sciverse.com/GadgetVideoPodcastPlayerWeb/verification} . This tool will display your video exactly in the same way as it will appear on ScienceDirect. }

\section{Impact}
% \textit{This is the main section of the article and reviewers will weight it appropriately.
% Please indicate:}
% \begin{itemize}
%     \item \textit{Any new research questions that can be pursued as a result of your software.}
%     \item \textit{In what way, and to what extent, your software improves the pursuit of existing research questions.}
%     \item \textit{Any ways in which your software has changed the daily practice of its users.}
%     \item \textit{How widespread the use of the software is within and outside the intended user group (downloads, number of users if your software is a service, citable publications, etc.).}
%     \item \textit{How the software is being used in commercial settings and/or how it has led to the creation of spin-off companies.}
%     \end{itemize}
% \textit{Please note that points 1 and 2 are best demonstrated by
%   references to citable publications.}

@@

\section*{Acknowledgements}
\label{}
Special thanks to Beni Altmann at The Comprehensive R Archive Network (CRAN) for help during the submission process to CRAN. We thank Tiffany Wolf and Marc Schwabenlander for their support through the Minnesota Center for Prion Research and Outreach. We would like to acknowledge Suzanne Stone and Sarah Gresch for maintaining lab operations.

@@

% \textit{If the software repository you used supplied a DOI or another
% Persistent IDentifier (PID), please add a reference for your software
% here. For more guidance on software citation, please see our guide for
% authors or \href{https://f1000research.com/articles/9-1257/v2}{this
%   article on the essentials of software citation by FORCE 11}, of
% which Elsevier is a member.}

% \large{\textbf{Reminder: Before you submit, please delete all 
% the instructions in this document, 
% including this paragraph. 
% Thank you!}}

>>>>

\documentclass[preprint,12pt,a4paper]{elsarticle}

@@

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
% \usepackage{lineno}
\usepackage{tabularray}
\usepackage{float}
\usepackage[T1]{fontenc}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\newcommand{\tab}[1]{\hspace{.2\textwidth}\rlap{#1}}

@@

        \title{quicR:\ An R Library for Streamlined Data Handling of Real-Time Quaking Induced Conversion Assays}
        \author[label1,label2,label3]{Gage Rowden\corref{cor}}
        \fntext[ ]{\textit{E-mail address: }rowde002@umn.edu}
        \cortext[cor]{Corresponding author.}
        \author[label1,label2,label3]{Peter Larsen}
        \address[label1]{Department of Veterinary and Biomedical Sciences, University of Minnesota, USA.}
        \address[label2]{Minnesota Center for Prion Research and Outreach, University of Minnesota, USA.}
        \address[label3]{Priogen Corp., USA.}

        \begin{abstract}
            Real-time quaking induced conversion (RT-QuIC) has become a valuable diagnostic tool for protein misfolding disorders such as Creutzfeldt-Jakob disease and Parkinson's disease. Given that the technology is relatively new, academic and industry standards for quality filtering data and high throughput analysis of results have yet to be fully established. The open source R library, quicR, was developed to provide a standardized approach to RT-QuIC data analysis.\ quicR provides functions, which can be easily integrated into existing R workflows, for data curation, analysis, and visualization.
        \end{abstract}

        \begin{keyword}
            R package \sep{} RT-QuIC \sep{} prion \sep{} diagnostics \sep{} CJD \sep{} Parkinson's
        \end{keyword}

@@

    \begin{table}[ht]
        \label{tbl:codeMetadata} 
        \begin{tabular}{|l|p{6cm}|p{6cm}|}
            \hline{}
            \textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata} \\
            \hline{}
            C1 & Current code version & V2.1.0 \\
            \hline{}
            C2 & Permanent link to code/repository used for this code version & \url{https://github.com/gage1145/quicR} \\
            \hline{}
            C3  & Permanent link to Reproducible Capsule & \url{https://cran.r-project.org/web/packages/quicR}\\
            \hline{}
            C4 & Legal Code License & GPL-3 \\
            \hline{}
            C5 & Code versioning system used & Git \\
            \hline{}
            C6 & Software code languages, tools, and services used & R \\
            \hline{}
            C7 & Compilation requirements, operating environments \& dependencies & R (>=4.1.0) \\
            \hline{}
            C8 & If available Link to developer documentation/manual & \url{https://cran.r-project.org/web/packages/quicR/quicR.pdf}\\
            \hline{}
            C9 & Support email for questions & rowde002@umn.edu\\
            \hline{}
        \end{tabular}
        \caption{Code metadata}
    \end{table}

@@

        By combining standardized metrics, advanced visualization tools, and a commitment to open source science, quicR serves as a foundational tool to empower researchers to analyze and present RT-QuIC data with clarity, consistency, and cohesion.

@@

    \section{Software description}
        \subsection{Software architecture}
            quicR was developed to address the growing need for efficient data conversion, analysis, and visualization of RT-QuIC data (Figure \ref{fig:workflow}). With a focus on usability and reproducibility, the package is designed to standardize workflows and ensure compatibility across multiple laboratories. 
            % Its primary input format is data exported as Excel workbooks from the proprietary MARS software (BMG Labtech, Ortenberg, Germany), providing seamless integration with existing experimental workflows.
            \begin{figure}[ht]
                \centering
                \includegraphics[width=\textwidth]{images/workflow.png}
                \caption{Workflow hierarchy of the quicR package. Blue nodes indicate steps where BMG software is needed. Purple nodes indicate functions dedicated to handling metadata. Red nodes are functions that acquire and manipulate raw data. Orange nodes are functions which calculate some metric. Finally, yellow nodes represent data analysis endpoints.}
                \label{fig:workflow}
            \end{figure}

    \subsection{Software functionalities}
        The implementation of the quicR package encompasses several streamlined processes designed to facilitate data input, cleaning, transformation, and analysis of RT-QuIC data. This section provides a comprehensive guide to utilizing the package's key functionalities, detailing how to:

        \begin{enumerate}
            \item Format and input sample data into Omega control software (BMG Labtech, Ortenberg, Germany).
            \item Extract, clean, and organize metadata and raw data and apply transformations and normalization for downstream analysis.
            \item Calculate critical analytical metrics, such as time-to-threshold (TtT), rate of amyloid formation (RAF), maxpoint ratio (MPR), and maximum slope (MS).
            \item Visualize raw and analyzed data.
        \end{enumerate}

        These steps are designed to enhance reproducibility, minimize manual data handling, and enable seamless integration with the MARS software (BMG Labtech, Ortenberg, Germany) workflow. Through practical examples, this section illustrates how each function operates, along with expected input and output formats, ensuring clarity and ease of use for researchers.

        \subsubsection{Input of Sample IDs into Omega Control Software}
            \begin{enumerate}
                \item \textbf{BMG\_format()}: The Omega control software allows input of a TXT file containing sample IDs, dilution factors, and their well locations. This file is uniquely formatted, and not easily reproduced manually. This function allows for input of a CSV file containing the plate layout (see Table \ref{tbl:layout} for proper formatting), and exports the formatted TXT file. The file can then be imported into the control software before running.
            \end{enumerate}

            \begin{table}[ht]
                \centering
                \label{tbl:layout}
                \begin{tblr}{
                    cells     = {font = \fontsize{11pt}{11pt}\selectfont},
                    colspec   = {|c|cccccccccccc|}, 
                    row{1}    = {font=\bfseries}, 
                    column{1} = {font=\bfseries}, 
                    rowhead   = 1,
                    width     = \textwidth
                }
                    \hline
                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\ 
                    \hline
                    A & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    B & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    C & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    D & P & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    E & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    F & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    G & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    H & N & S01 & S02 & S03 & S04 & S05 & S06 & S07 & S08 & S09 & S10 & S11 \\ 
                    \hline
                \end{tblr}
                \caption{Example CSV file plate layout for input into the ``BMG\_format'' function. The top left corner should be cell ``A1'' in the CSV file. The top numbered row and the left-most lettered column should never be altered.}
            \end{table}

        \subsubsection{Data Cleaning and Transformation}
            The MARS software exports real-time data as an Excel workbook. Typically, the first sheet in the workbook will include microplate views of both raw data and metadata; however, the metadata on this page is what is most useful for downstream processes. Those tables include the ``Sample IDa'' and ``Dilutions'' tables (if dilutions were included in the MARS export). For much of the downstream analysis, it is crucial that the ``Sample IDs'' table was exported from MARS.\@ If there is no table, the user can simply add it manually (see [@fig-sheet1] for proper formatting).

            \begin{enumerate}
                \item \textbf{organize\_tables()}: returns a list of tables contained in the first sheet of the exported Excel sheet. These tables contain valuable metadata such as sample IDs, dilution factors, and microplate locations.
                \item \textbf{convert\_tables()}: accepts tables outputted from \textbf{organize\_tables} and converts them to columns in a data frame.
                \item \textbf{get\_sample\_locations()}: extracts the well locations for each sample. Output of this function is used as an argument for visualizing a microplate-level view of real-time data.
            \end{enumerate}

        \subsubsection{Retrieving and Manipulating Raw Data}
            The raw, real-time data is typically found on the second sheet of the Excel workbook exported from MARS. There are three functions dedicated to the retrieval and cleaning of raw data.

            \begin{enumerate}
                \item \textbf{get\_real()}: Retrieves the raw data from the Excel file, and outputs it as a data frame.
                \item \textbf{transpose\_real()}: Swaps the rows and columns which makes some downstream analyses easier.
                \item \textbf{normalize\_RFU()}: normalizes the raw data by dividing each read by background fluorescence at a given cycle.
            \end{enumerate}

        \subsubsection{Calculations}
            Three analytical metrics have dedicated functions: time-to-threshold (TtT), maxpoint ratio (MPR), and maximum slope (MS). The rate of amyloid formation does not have a designated function since it is simply the reciprocal of the time-to-threshold (1/TtT). Each function below accepts input from the transpose\_real() or the normalize\_RFU() functions. See Figure \ref{fig:metrics} for an example of the output of these functions.

            \begin{enumerate}
                \item \textbf{calculate\_threhold()}: returns a value which is a given number of standard deviations above the average background fluorescence of the entire microplate.
                \item \textbf{calculate\_TtT()}: takes the real-time data and calculates the time in hours needed to reach a given threshold value.
                \item \textbf{calculate\_MPR()}: accepts raw or normalized data and returns the maximum value obtained during the run. If supplied with raw data, it will make a call to the normalize\_RFU() function.
                \item \textbf{calculate\_MS()}: computes the approximate derivative of the real-time data and returns the maximum value obtained during the run.
            \end{enumerate}

            \begin{figure}[ht]
                \centering{}
                \includegraphics[width=\textwidth]{images/metric_example.png}
                \caption{Example graph highlighting the calculated metrics described above. The red curve represents a raw data curve that has been normalized against background. The maxpoint ratio is calculated as the maximum fluorescent value achieved in the normalized raw data. Time‐to‐threshold is determined as the time required to cross a given threshold (in this example, the threshold is set at 0.2). The blue curve represents the derivative of the raw data, and max slope is determined as the maximum of the derivative.}
                \label{fig:metrics}
            \end{figure}

        \subsubsection{Visualization}
            The ensuing goal of this package is visualizing RT-QuIC data. While there are many ways to represent this data, quicR includes two functions which provide a quick assessment of the reactions.
            \begin{enumerate}
                \item \textbf{plate\_view()}: accepts the output from the get\_real() function and get\_sample\_locations() function. Makes an 8$\times$12 or 16$\times$24 faceted plot depending on if the plate has 96 or 384 wells. Each facet shows the real-time data of each well.
                \item \textbf{plot\_metrics()}: generates a boxplot of each sample's MPR, MS, RAF, and TtT.
            \end{enumerate}

        % \subsection{Sample code snippets analysis (optional)}
    \section{Illustrative examples}

@@

    % \textit{\textbf{Optional}: you may include one explanatory  video or screencast that will appear next to your article, in the right hand side panel. Please upload any video as a single supplementary file with your article. Only one MP4 formatted, with 150MB maximum size, video is possible per article. Recommended video dimensions are 640 x 480 at a maximum of 30 frames / second. Prior to submission please test and validate your .mp4 file at  \url{http://elsevier-apps.sciverse.com/GadgetVideoPodcastPlayerWeb/verification} . This tool will display your video exactly in the same way as it will appear on ScienceDirect. }
        To demonstrate the utility of quicR, we used a typical RT-QuIC run performed on a 96-well plate. The reaction was performed on a FLUOstar Omega plate reader (BMG Labtech, Ortenberg, Germany). The file was exported as an Excel worksheet where the metadata appears on the first sheet and the raw data on the second sheet.

        \subsection{Example Code}
            \# Step 1: Identify the raw file.\\
            file <- "example.xlsx"\\

            \# Step 2: Extract the raw data.\\
            raw <- get\_real(file)[[1]]\\

            \# Step 3: Normalize the data against the background.\\
            normal <- normalize\_RFU(raw, transposed = FALSE)\\

            \# Step 4: Extract the metadata.\\
            meta <- file |> organize\_tables() |> convert\_tables()\\

            \# Step 5: Get sample locations.\\
            locations <- get\_sample\_locations(file)\\

            \# Step 6: Create the analyzed data frame.\\
            analyzed <- calculate\_metrics(normal, meta, dilution\_bool = TRUE)\\

            \# Step 7: Plot the analyzed data frame.\\
            plot\_metrics(analyzed)\\

            \# Step 8: Plot the plate view.\\
            plate\_view(raw, locations)\\

            Once the file name is identified, the function get\_real() is used to extract the raw data. This data is imported as a data frame with each sample designated as an individual column and the first column designated as time. Much of the downstream analysis works best with background-normalized data, so the raw data is then passed to the normalize\_RFU() function. This function accepts an integer as an argument for the desired cycle to be selected for background measurement. Next, the metadata is extracted using the organize\_tables() and convert\_tables() functions. These take the metadata found on the first sheet of the Excel file and convert them into data frame columns. Finally, metrics are calculated and assigned to their given sample ID using calculate\_metrics(). This generates a data frame with sample IDs, dilution factors (if chosen), MPR's, MS's, RAF's, and TtT's.

            Once the proper variables have been defined, they can be used as input in the visualization functions. The analyzed data frame can be used as an argument in plot\_metrics() to generate a plot such as \ref{fig:boxplot}. Additionally, the raw data and sample locations are used to plot the plate view as in \ref{fig:plateview}.

            \begin{figure}[ht]
                \centering
                \includegraphics[width=\textwidth]{images/boxplot.png}
                \caption{Boxplot of the four metrics calculated in the quicR package.}
                \label{fig:boxplot}
            \end{figure}

            \begin{figure}[ht]
                \centering
                \includegraphics[width=\textwidth]{images/plate_view.png}
                \caption{Plate view of RT-QuIC. Wells are identified with a sample ID and dilution factor separated by an optional delimiter (in this case, a whitespace).}
                \label{fig:plateview}
            \end{figure}

    \section{Impact}

        The quicR package represents a significant advancement in the standardization and reproducibility of RT-QuIC data analysis. By integrating key metrics such as time-to-threshold (TtT) \cite{Orru2015}, maxpoint ratio (MPR) \cite{Rowden2023}, and maximum slope (MS) \cite{Henderson2015}, quicR addresses critical gaps in the field, providing researchers and diagnosticians with a robust toolkit for interpreting RT-QuIC data.

        One of the primary strengths of quicR lies in its flexibility and user-centric design. The package leverages R’s powerful ecosystem, including the tidyverse and ggplot2, to streamline workflows and create high-quality, customizable visualizations. This ensures accessibility for a wide range of users, from experienced data scientists to wet-lab researchers unfamiliar with programming. Additionally, the incorporation of open-source principles allows the broader scientific community to contribute to its development, fostering innovation and adaptability.

        Despite these strengths, there are limitations to consider. Currently, quicR is tailored to data exported from the MARS software, which may limit its applicability to researchers using alternative microplate readers. Future iterations of the package could expand compatibility by incorporating functions to handle diverse data formats. Furthermore, while quicR includes robust visualization tools, users seeking highly specialized plots may require additional customization beyond the package’s default capabilities.

        Another avenue for improvement lies in the standardization of RT-QuIC diagnostic criteria. quicR provides tools to calculate key metrics, but consensus on thresholds and interpretations remains a challenge for the field \cite{Rowden2023}. Collaborative efforts among researchers and clinicians are necessary to define universal criteria, enabling quicR to fully realize its potential as a diagnostic aid. Diagnostic determinations could easily be built into the library, but a larger consensus within the research community will need to be reached to warrant inclusion.

@@

    \section*{Acknowledgements}
        Special thanks to Beni Altmann at The Comprehensive R Archive Network (CRAN) for help during the submission process to CRAN. We thank Tiffany Wolf and Marc Schwabenlander for their support through the Minnesota Center for Prion Research and Outreach. We would like to acknowledge Suzanne Stone and Sarah Gresch for maintaining lab operations.

@@

    \textit{If the software repository you used supplied a DOI or another
    Persistent IDentifier (PID), please add a reference for your software
    here. For more guidance on software citation, please see our guide for
    authors or \href{https://f1000research.com/articles/9-1257/v2}{this
      article on the essentials of software citation by FORCE 11}, of
    which Elsevier is a member.}
